---
title: "Reproducible Research Assignment 1"
author: "Aaron Augustine"
date: "Tuesday, June 09, 2015"
output:
  html_document: default
---
 
This document outlines the results for Reproducible Research Assignment 1.

### Loading and preprocessing the data

Show any code that is needed to

1. Load the data (i.e. `read.csv()`)

```{r, echo=FALSE, warning=FALSE}
#reference packages
library(downloader)
library("plyr")
library(lattice) 
```


```{r}
filename <-paste(getwd(),"/activity.zip",sep="")  #set file name to unzip
unzip (filename, exdir = getwd())                 #unzip file for analysis


### Loading and preprocessing the data
rawdata<-read.csv("./activity.csv",
                  header=TRUE,
                  sep=",",
                  colClasses=c("numeric","character","numeric"),
                  na.strings = "NA"
)                                                 #read in raw .csv file      
```

2. Process/transform the data (if necessary) into a format suitable for your analysis

```{r}
str(rawdata)                                      #check formats
rawdata$date <- as.Date(rawdata$date)             #convert charcater date to actual date format
str(rawdata)                                      #check formats     
```

### What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in
the dataset.

1. Make a histogram of the total number of steps taken each day
```{r}
totalsteps<-ddply(rawdata,c("date"),
                  summarize,
                  totalsteps=sum(steps,
                                 na.rm=TRUE))     #calculate total number of steps by day

summary(totalsteps$totalsteps)
hist(totalsteps$totalsteps,
     xlab="Total Steps by Day",
     ylab="Frequency",
     ylim=c(0,40),
     main="Histogram of Total Steps by Day"
)                                                 #histogram of total steps by day

```

2. Calculate and report the **mean** and **median** total number of steps taken per day

```{r}
meansteps<-mean(totalsteps$totalsteps,na.rm=TRUE) #calculate mean total steps
meansteps

mediansteps<-median(totalsteps$totalsteps,na.rm=TRUE) # calcualte median total steps
mediansteps
```

### What is the average daily activity pattern?

1. Make a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
avgsteps<-ddply(rawdata,c("interval"),summarize,
                  avgsteps=mean(steps,
                  na.rm=TRUE))     #calculate mean number of steps by day

with(avgsteps, plot(interval, avgsteps
                   ,xlab="Interval"
                   ,ylab="Avg Steps across Days"
                   ,type="l"))
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
maxsteps<-avgsteps[avgsteps$avgsteps==max(avgsteps$avgsteps,na.rm=TRUE),]
maxsteps   
```

### Imputing missing values

Note that there are a number of days/intervals where there are missing
values (coded as `NA`). The presence of missing days may introduce
bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with `NA`s)
```{r}
nummissing<-sum(is.na(rawdata$steps))             #number of missing records in the file
nummissing
```
2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r}
```
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
adjdata <- merge(rawdata,avgsteps,
                 by.x="interval",
                 by.y="interval",
                 all.x=TRUE)
adjdata$adjsteps <- adjdata$steps                 #copy steps into adjsteps
adjdata$adjsteps[is.na(adjdata$steps)] <- adjdata$avgsteps[is.na(adjdata$steps)] 

```
4. Make a histogram of the total number of steps taken each day and Calculate and report the **mean** and **median** total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
adjtotalsteps<-ddply(adjdata,c("date"),
                  summarize,
                  totalsteps=sum(adjsteps,
                                 na.rm=TRUE))     #calculate total number of steps by day

summary(adjtotalsteps$totalsteps)
hist(adjtotalsteps$totalsteps,
     xlab="Total Steps by Day",
     ylab="Frequency",
     ylim=c(0,40),
     main="Histogram of Total Steps by Day adjusted for NAs"
)                                                 #histogram of total steps by day


adjmeansteps<-mean(adjtotalsteps$totalsteps,na.rm=TRUE) #calculate mean total steps
adjmeansteps

adjmediansteps<-median(adjtotalsteps$totalsteps,na.rm=TRUE) # calcualte median total steps
adjmediansteps

```

### Are there differences in activity patterns between weekdays and weekends?

For this part the `weekdays()` function may be of some help here. Use
the dataset with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
adjdata$weekday <- weekdays(adjdata$date)
```
1. Make a panel plot containing a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was created using **simulated data**:
```{r}
adjdata<-transform(adjdata ,
                   weekend = ifelse(weekday %in% c("Sunday","Saturday"),
                                    "Weekend",
                                    "Weekday" ))
avgstepsbyweekend<-ddply(adjdata,c("interval","weekend"),summarize,
                         avgsteps=mean(steps,
                                       na.rm=TRUE)) 
                                        #calculate mean number of steps by weekend
xyplot(avgsteps~interval | weekend, data = avgstepsbyweekend,
       type = 'l',
       xlab = 'Interval',
       ylab = 'Number of Steps',
       main = 'Number of Steps across intervals, Weekends vs. Weekdays',
       layout = c(1,2))

```